---
title: "Project Proposal : Stats 290 : Minakshi Mukherjee"
output:
     html_document
---

__Proposal Deadline: Feb 3, 2019 23:59:00 hrs.__

- Edit this markdown file by __replacing only__ what is there in the
chunks named `q0`, `q1`, `q2`, ..., `q5` with your team's appropriate
answers.

- Upload the filled in markdown to the project proposal assignment
area (will be created shortly) in every one of the team member's
private directory with the name `proposal.Rmd`.

0. Who are the authors of this package? See format below where you
   should replace `sunet1`, `sunet2`, and `Full Name1` and `Full
   Name2`.

```{r q0}
authors <- list(adaboost = "Minakshi Mukherjee")
```

1. What is your package trying to achieve?  Who is it aimed at?

```q1

1.  Querying.  The package will enable scientists and users to perform exploratory 
data analysis for Light Detection and Ranging(LiDAR) data. Usually each LiDAR 
file is very large, so it is computationally intensive to address scientific questions. 
This package will provide step by step approach to perform statistical analysis on 
LiDAR dataset and will provide classification techniques to distinguish among 
vegetation, buildings and dry land. The package will use Open-Source embedded 
database MonetDBLite. This columnar database will showcase few analytic queries 
and bechmark those against base R query output.

*** This outline has been mostly followed in the project implementation.

2.  Visualization.  LiDAR dataset consists of billions of points, where each point has a 
X, Y, Z value and other attributes. LIDAR is a very accurate and precise technology that 
uses laser pulse to strike an object, so it captures the surface elevation usually 
hidden by vegetation or forest canopy. In order to turn billions of LiDAR points into 
Digital Surface Models(DSMs), Digital Terrain Models(DTMs), Canopy Height Models(CHMs) 
and elevation contours, this package will use the C++ package LAStools as the fastest 
and most memory efficient mechanism for LiDAR processing. 

*** Due to the limited time frame of the project, C++ package LAStools has not been used
in the project implementation, instead all functions have been developed using R.

3.  Analysis.  This package will include multiple functions for the following tasks:
> Read LiDAR files and print summary

*** Implemented: readFile.R, lasSlotExtract.R, lasSlotPrint.R, processTiff.R

> Use LAStools to explore Digital Surface Models, Digital Terrain Models, 
Canopy Height Models, several elevation contours etc

*** Instead of C++ package LAStools, R function has been developed.

Implemented: lasVisualize.R

> Use LAstools to prepare a meaningful dataset that can be fed into different
classification algorithms in R

*** In addition to the above R functions, treeDensityIndex.R and treeDensityPlot.R 
are implemented to prepare a meaningful dataset.

> Use optimized Bagging algorithm to perform classification 
of vegation, building and dry land.

*** A typical LiDAR file comes as several Gigabytes in size. In order to understand a 
topoloy or to proceed with vegetation segmentation, feature extraction etc, 100k-200k 
files of this size need to be processed which requires a cloud computing environment.
Due to the limited time frame and lack of proper computing resources, the classification 
component is eliminated from the project.

> Showcase open source columnar database MonetDB to perform query analysis. 

*** loadMonetDB.R showcases a few database queries.
```

2. Briefly describe the background if that's not obvious.  What field
   or area of interest is it related to?  (Just briefly, in a paragraph
   or two, maximum.) References, including web references, are welcome
   where available.

```q2

This package design is inspired by the R&D effort of my organization who is trying to 
analyze and understand LiDAR data in the context of how close the power line is to 
the vegetaion. At present, it is an open problem in our company. My objective is to 
build a comprehensive LiDAR data analysis package to illustrate step by step data 
exploration and classification. The visualization and report from this package will 
equip other teams to proceed to the next phase of utilizing LiDAR data.

https://www.informatica.com/content/dam/informatica-com/global/amer/us/collateral/white-paper/grid-health-360_white-paper_3487en.pdf

Apart from the above use case, LiDAR data has a lot of other use cases. 
Since LiDAR dataset always comes in a standard vector format, my package can 
be used as an exploratory data analysis and visualization tool for almost 
every use cases. Several R packages for LiDAR have been built over a period of 
time, however there is always a need for a single package that can show all 
different types of LiDAR data exploration. Also, there are always further 
opportunities to incorporate more C++ packages into LiDAR analysis to speed 
up processing.

Some of the other important use cases of LiDAR are: Agriculture, Forest Fire 
Management, Tourism and Parks management, Biodiversity of Birds, Building 3D 
simulation of River Bank and Flood Model, Transport Planning, CoastLine 
Management, Archeology, Glacier Volume Change monitoring, Use of Differential 
Absorption LiDAR(DIAL) to trace amount of gases above the hydrocarbon 
region and so on..
```

3. Provide one or two use cases; that is, in a sentence or so say how
   you see someone using the package.

```q3

Scientists or users will pass their LiDAR files to the R package and 
it will produce a visualization for surface model, terrain model, 
canopy height etc..

Users will have options to extract elevation data to be used for their individual 
use cases. This data will be stored into an open source columnar database MonetDB
to show sample queries.

There will be examples and test cases showing an optimized bagging algorithm.
Classification algorithm will be trained on the given dataset and prediction 
output will be displayed on a different test data set. Users will get idea as 
how to train this types of dataset and what parameters to pass. 

*** Vignette shows a clear explanation of all the use cases of the package.
```

4. If you can, list the main functions or other software you plan on
   including.  No need to implement anything now, but you should have
   an idea.

```q4

The R package will provide the following functions: reading .las or 
.laz files for LiDAR data and printing summary; creating DSM, DEM, 
DTM, CHM models using LAStools to produce graphs and visualizations;
preparing a meaningful and simplified dataset out of the original LiDAR 
files as an input to train classification model using optimized 
bagging algorithm; show predictions on a test dataset using the 
trained model.
The R package will also provide MonetDB sample queries and outputs 
for deeper analysis by the user.

*** Implemented all of above in these R functions:
readFile.R
processTiff.R
lasSlotExtract.R
lasVisualize.R
loadMonetDB.R
treeDensityIndex.R
treeDensityPlot.R
```

```{r, echo = FALSE, hide = FALSE}
## Do not edit
none <- "none"
basic <- "basic"
proficient <- "proficient"
dangerous <- "dangerous :-)"
```

5. For languages C/C++, Java, Python, Ruby, please specify your
   proficiency: Proficient, basic knowledge, no knowledge.  (This
   helps us evaluate the proposal sometimes and offer suggestions.)

```{r q5}
languages <- list(
    sunet1 = c(cxx = basic,
               java = none,
               python = proficient,
               julia = none, 
               ruby = none)
)
```

## Comments and Responses

__Note__ Your responses should be in this markdown itself, right after
each point. You do not need to submit or email the responses, but do
update it on your github repo exactly where you found it.

1. We discussed your project in person. So you already know of some
   comments.

2. I think this can be a large project, so do identify things you can
   do well: data handling, visualization.  The machine learning part
   can take time and so I suggest you do that last.













